---
output:
  html_document:
    keep_md: yes
---
# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

  1. Load the data (i.e. read.csv() )
  2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
## Read the csv file from the activity.zip file.
## using unz function to load a zip file instance and read the activity.csv file stored in
## the zip file.
activity <- read.csv(unz("activity.zip", filename="activity.csv"),quote="\"", as.is = TRUE)

## Data Processing, First convert the date variable in string format to Date object.
activity$date <- as.Date(activity$date,"%Y-%m-%d")
activity_no_missing <- activity[!is.na(activity$steps),]
```

## What is mean total number of steps taken per day?

  For this part of the assignment, you can ignore the missing values in the dataset.
  1. Make a histogram of the total number of steps taken each day

```{r}
## Factor data by date.
date_factor = factor(activity_no_missing$date)

## Calculate the sum of steps for each day
steps_day_total <- tapply(activity_no_missing$steps, date_factor,sum)

## Plot the Histogram
hist(steps_day_total, main = 'Total Number Of Steps Taken Each Day', xlab='Number of Steps')
```

  2. Calculate and report the mean and median total number of steps taken per day

```{r showtable,results='asis'}
# Calculate mean of steps for each day
steps_day_mean <- tapply(activity_no_missing$steps, date_factor,mean)

# Calculate median of steps for each day
steps_day_median <- tapply(activity_no_missing$steps, date_factor,median)

# Create a data frame containg mean and median data for each day 
table_data <- data.frame(day_mean = steps_day_mean, day_median = steps_day_median)

# Load the xtable library to render data in tabular format.
library(xtable)

# Create xtable variable
xtable <- xtable(table_data,label="1")

# Render xtable as html table
print(xtable, type ="html")
```

## What is the average daily activity pattern?

  1. Make a time series plot (i.e. type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
interval_factor <- factor(activity_no_missing$interval)
steps_interval_mean <- tapply(activity_no_missing$steps, interval_factor,mean)
plot(names(steps_interval_mean), steps_interval_mean, ylab = "Average # of Steps Taken", xlab = "Interval", type = "l")
```

  2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
# Subset steps_interval_mean by max(steps_interval_mean) and
# get the names to fetch the interval.
max_mean <- max(steps_interval_mean)
max_interval <- names(steps_interval_mean[max_mean == steps_interval_mean])
```
Maximum mean *`r max_mean`* at interval *`r max_interval`*

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA's)
```{r}
  missing_count <- nrow(activity) - nrow(activity_no_missing)
```
Total number of missing values are: `r missing_count` rows

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

  ***We will use the mean for the 5-minute Interval to replace missing values***
  
  1. Use the *steps_interval_mean* list computed above. 
  2. Identify missing data rows in *activity* data frame.
  3. Replace *activity$step* having NA with *steps_interval_mean* value for a given interval.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

  ***Using strategy mentioned in previous step create activity_fixed_missing***
```{r}
# First create a copy of original data frame
# activity_fixed_missing <- activity was throwing a strange error in knit-er environment
# on the date variable.
#
# For the purpose of meeting assessment end-date, using data.frame class to create a 
# copy of activity (again for some strange reasons I need to convert date to string object)
# and converting the date variable back to Date object.
# This is working for me now, again I will explore the issue later for better approach.
activity_fixed_missing <- data.frame(steps = activity$steps, date = as.character(activity$date,"%Y-%m-%d"), interval = activity$interval)
activity_fixed_missing$date <- as.Date(activity_fixed_missing$date,"%Y-%m-%d")

# Again I am using very primitive approach to replace the missing value in for loop
# I am sure there is a better approach using apply function that I need to explore.
# But the code below works.
for(i in c(1:nrow(activity_fixed_missing))) {
  if(is.na(activity_fixed_missing$steps[i])) 
    activity_fixed_missing$steps[i] = 
      steps_interval_mean[names(steps_interval_mean) == activity_fixed_missing$interval[i]]
}
summary(activity_fixed_missing)
```
  ***Note: Note the summary above shows there is no missing values***

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

  ***Histogram of total number of steps taken each day for fixed data***
```{r}
## Factor data by date.
date_factor_fixed = factor(activity_fixed_missing$date)
## Calculate the sum of steps for each day
steps_day_total_fixed <- tapply(activity_fixed_missing$steps, date_factor_fixed,sum)
## Plot the Histogram
hist(steps_day_total_fixed, main = 'Total Number Of Steps Taken Each Day', xlab='Number of Steps')
```

  ***Mean and Median for fixed data***
```{r showtable2,results='asis'}
steps_day_mean2 <- tapply(activity_fixed_missing$steps, date_factor_fixed,mean)
steps_day_median2 <- tapply(activity_fixed_missing$steps, date_factor_fixed,median)
table_data2 <- data.frame(day_mean = steps_day_mean, day_median = steps_day_median)

xtable <- xtable(table_data2,label="1")
print(xtable, type ="html")
```


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating
whether a given date is a weekday or weekend day.

```{r}

## Factor data by Weekdays and Weekend, considering Saturday and Sunday as Weekend.
activity_fixed_missing$weekdays = ifelse(weekdays(activity_fixed_missing$date) %in% c("Saturday","Sunday"), "Weekend", "Weekday")
factor_weekdays <- factor(activity_fixed_missing$weekdays)

```
2. Make a panel plot containing a  time series plot (i.e. type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:
```{r}
## Using lattice to plot the graph
## For this to work install lattice using install.packages()
library(lattice)
library(plyr)
## Factor data by Weekdays and Weekend, considering Saturday and Sunday as Weekend.
factor_interval <- factor(activity_fixed_missing$interval)

##Create a Dataframe by grouping them by weekdays and iterval and calculating mean of each group
DayIntervalAverage = ddply(activity_fixed_missing, .(interval,weekdays), function(x,...) {mean(x$steps,rm.na=TRUE)})

##Refactor weekdays for this new dataframe
factor_weekdays <- factor(DayIntervalAverage$weekdays)

## Plot the graph
xyplot(DayIntervalAverage[,3] ~ DayIntervalAverage$interval | factor_weekdays, layout = c(1, 2), type="l", ylab = "Number of Steps", xlab="Interval")

```

End of Report
